Python as a machine learning language: experiences from the scikit-learn project

### 2013-04-26 api.tex
expand/restructure API section

The bulk of the scikit-learn application programming interface
consists of so-called \textit{estimator} objects.
All learning algorithms for classification, regression and clustering
are offered as such objects,
but so are feature extraction, feature selection and dimensionality reduction.
What all estimators have in common is a method called \texttt{fit},
which causes the estimator to learn from data supplied as an argument.
Such a method may seem superfluous for tasks such as vectorizing text documents
for subsequent learning,
but even the classes that handle this task can be said
to learn their vocabulary from the training corpus.
(Only in a very small number of cases,
specifically two estimators that implement the ``hashing trick'' of
\citealt{weinberger2009} for feature extraction,
is \texttt{fit} a no-op offered only for consistency and composibility.)

The \texttt{fit} method's task is to learn model-specific parameters
from a training set and set these as attributes on the estimator object,
which can then be used to transform data or make predictions.
As in other object-oriented machine learning libraries
(e.g.\ Weka's Java API, \citealp{hall2009weka})
no distinction is made in the type hierarchy
between what is variously called an estimator or a \textit{learner}
in the machine learning literature,
and the \textit{model} that results from applying an estimator to data.
When the \texttt{fit} method has been called,
an estimator object serves dual purpose as a model
(and is referred to as such in the documentation).

Importantly, the parameters learned by an estimator
are exposed as \textit{public} attributes on the trained object.\footnote{
  The Python language does not actually distinguish between public and private
  attributes and methods,
  but by convention, any identifier prefixed with an underscore (\texttt{\_})
  is considered ``a non-public part of the API\ldots
  an implementation detail and subject to change without notice.''}
This facilitates model inspection
and makes it possible to train a model using scikit-learn,
export it to a (standardized or custom) external data format
and use the learned parameters in a different piece of code,
perhaps a prediction algorithm implemented in a different language
or one that combines the model parameters with other information
to build a custom model.
Of course, it also forces the developers to think carefully
about what might in a more ``black box''-style toolkit
be considered the internals of objects,
since authors are client code are allowed to rely
on the names and formats of model parameters.

% subsection here? also discuss scipy.sparse?
Input to an estimator is typically given as NumPy arrays \citep{vanderwalt2011};
typically, a clustering or dimensionality reduction algorithm
takes a single NumPy array, conventionally called $X$ of size $n \times m$
to represent $n$ samples of $m$ features each.
A regression or classification estimator takes, at training time,
an additional array of size $n$, by convention called $y$,
that represents the target variable.
While the NumPy array may seem like a ``bare'' representation of data
when compared to more object-oriented representations
(again, Weka is an example of this style of API),
it brings the prime advantages of allowing \textit{vector operations}
on entire arrays,
which are often orders of magnitude faster than the corresponding Python loops
while at the same time keeping the code short and readable.
Although conversion to the NumPy format may seem like a burden on the user,
we argue that this is not the case.
Python's lists (its dynamic array structure)
are easily transformed into NumPy arrays
and many scientific users of Python will already be handling such arrays,
since they pervade in other science-oriented Python packages.
For tasks where input is likely to consists of text files
or semi-structured objects, ``vectorizers'' are provided
that efficiently convert from such data formats to the NumPy format.
Finally, offering a custom datatype for samples
would require a conversion \textit{in any case}.

@@

A distinguishing feature of scikit-learn estimators
is that they can be composed into \texttt{Pipeline} objects.
Not only do such objects combine typical machine learning workflows
(extract features, center/scale/normalize, select features, 
make predictions) in a single object that is itself an estimator
whose methods cause data to flow through the pipeline,
but more importantly they can be used for
systematic, automated model selection.
This is best explained by an example.

>>>>

\subsection{Estimators}

% TODO describe transformer vs. classifiers/regressors

The bulk of the scikit-learn application programming interface
consists of so-called \textit{estimator} objects.
All learning algorithms for classification, regression and clustering
are offered as such objects,
but so are feature extraction, feature selection and dimensionality reduction.

Construction and application of an estimator object
proceeds in a way that is somewhat reminiscent of partial function application.
The constructor, \texttt{\_\_init\_\_}\footnote{
  Strictly speaking, \texttt{\_\_init\_\_} is Python's
  class \textit{initialization} method,
  while \texttt{\_\_new\_\_} is the constructor.
  The distinction is not relevant to the present description of scikit-learn.}
takes a set of named (hyper-)parameters
and attaches these to the estimator as public attributes.\footnote{
  The Python language does not actually distinguish between public and private
  attributes and methods,
  but by convention, any identifier prefixed with an underscore (\texttt{\_})
  is considered ``a non-public part of the API\ldots
  an implementation detail and subject to change without notice.''
  % TODO cite Python tutorial here
  }
It never performs actual learning, nor does it even see a training set.

The workhorse of most estimators, and one of the few methods
that all estimators have in common, is called \texttt{fit}.
Calling this method causes the estimator to learn from a training set,
supplied as one or more arguments.
Its task, from an API point of view,
is to determine (via a learning algorithm) model-specific parameters
from a training set and set these as attributes on the estimator object,
which can then be used to transform data or make predictions.
As in other object-oriented machine learning libraries
(e.g.\ Weka's Java API, \citealp{hall2009weka})
no distinction is made in the type hierarchy
between what is variously called an estimator or a \textit{learner}
in the machine learning literature,
and the \textit{model} that results from applying an estimator to data.
When the \texttt{fit} method has been called,
an estimator object serves dual purpose as a model
(and is referred to as such in the documentation).

Importantly, the parameters learned by an estimator
are exposed as \textit{public} attributes on the trained object.
This facilitates model inspection
and makes it possible to train a model using scikit-learn,
export it to a (standardized or custom) external data format
and use the learned parameters in a different piece of code,
perhaps a prediction algorithm implemented in a different language
or one that combines the model parameters with other information
to build a custom model.
Of course, it also forces the developers to think carefully
about what might in a more ``black box''-style toolkit
be considered the internals of objects,
since authors are client code are allowed to rely
on the names and formats of model parameters.

A \texttt{fit} method is present even on objects
that perform seemingly mundane tasks such as vectorizing text documents
for subsequent learning.
Such classes, however, can be said to learn their vocabulary
from the training corpus,
and follow the conventions set out above.
Even those estimators that are entirely stateless and therefore
do not require a \texttt{fit} method to perform useful work,
nevertheless have such a method for consistency and composibility,
and to perform parameter validation.
Examples of this latter kind of objects are the normalizing transformer,
kernel approximation transformers,
and two feature extraction classes that implement the ``hashing trick''
\citep{weinberger2009}.

\subsection{Data formats}

% TODO also discuss scipy.sparse
Input to an estimator is typically given as NumPy arrays \citep{vanderwalt2011};
typically, a clustering or dimensionality reduction algorithm
takes a single NumPy array, conventionally called $X$ of size $n \times d$
to represent $n$ samples of $d$ features each.
A regression or classification estimator takes, at training time,
an additional array of size $n$, by convention called $y$,
that represents the target variable.
While the NumPy array may seem like a ``bare'' representation of data
when compared to more object-oriented representations
(again, Weka is an example of this style of API),
it brings the prime advantages of allowing \textit{vector operations}
on entire arrays,
which are often orders of magnitude faster than the corresponding Python loops
while at the same time keeping the code short and readable.
Although conversion to the NumPy format may seem like a burden on the user,
we argue that this is not the case.
Python's lists (its dynamic array structure)
are easily transformed into NumPy arrays
and many scientific users of Python will already be handling such arrays,
since they pervade in other science-oriented Python packages.
For tasks where input is likely to consists of text files
or semi-structured objects, ``vectorizers'' are provided
that efficiently convert from such data formats to the NumPy format.
Finally, offering a custom datatype for samples
would require a conversion \textit{in any case}.

@@

A distinguishing feature of scikit-learn estimators
is that they can be composed into \texttt{Pipeline} objects.
Not only do such objects combine typical machine learning workflows
(extracting features, centering/scaling/normalization,
feature selection, making predictions)
in a single object that is itself an estimator,
and whose methods cause data to flow through the pipeline.
More importantly, though, they can be used for
systematic, automated model selection.
This is best explained by example.
